Why do languages tolerate heterography? An experimental investigation of the cultural evolution of informative orthography

### 2023-08-30 manuscript/wp4.tex
ms 30/08

Fig.~\ref{cost} plots communicative cost over the nine generations of cultural transmission for both the Transmission-only condition (left) and Transmission + Communication condition (right). Each colored line represents an individual chain, and the thick black lines shows the mean across chains. The initial, generation-0 seed systems are all fully informative (zero cost) because each meaning is signaled by a unique arbitrary spelling. In both conditions, cost increases with generation, implying that the languages become less informative over time. This is to be expected in the Transmission-only condition, since there is only pressure from learnability, so the systems evolve to become simple and transparent. In contrast, this is not the expected result in Transmission + Communication, where the communicative pressure should promote the emergence of informative systems, but it does align with our qualitative interpretation of the data, which suggested that participants in the communicative condition rarely produce differentiated spellings, be it through explicit innovation or implicit generalization. Note, however, that the cost plateaued at around 1.3~bits in Transmission-only and around 1~bit in Transmission + Communication. This suggests that there is some tendency for the systems to remain more informative in the communicative condition.

To evaluate this argument statistically, we regressed cost on generation in a Bayesian hierarchical cubic regression model with a gamma likelihood and log link function. The posterior parameter estimates are shown in Fig.~\ref{stats1}. Essentially, the beta coefficients of the model (representing the shape of the polynomial) do not differ between conditions, but the intercepts do differ (generation was centered, such that the intercept corresponds to generation 5). This difference in intercepts reflects the fact that the cost plateaued at a lower level in the Transmission + Communication condition. Nevertheless, the posteriors for $\alpha_\mathrm{tran}$ and $\alpha_\mathrm{comm}$ do overlap to some extent, suggesting no definitive difference between conditions.

@@

As in Experiment~1, we plot communicative cost by generation in both conditions (see Fig.~\ref{cost2}). Each colored line represents an individual chain, and the thick black lines shows the mean across chains. These mean lines reveal a fairly clear effect from epoch. During Epoch~I, there is a shallow increase in cost with generation in both conditions. This represents a kind of baseline level of noise when the spoken languages are fully informative. During Epoch~II, there is a steep rise in cost in the Transmission-only condition, but a relatively shallow rise in Transmission + Communication. This is because the communicative pressure acts to slow down the rate at which the systems become transparent or redundant. In Epoch~III, the cost curve plateaus in the Transmission-only condition because the written languages have become sufficiently simple to be reliably learned and reproduced with little change. In Transmission + Communication, by contrast, the curve gets a little steeper because the systems still have some room to simplify and the pressure to simplify has been ramped up.

@@

>>>>

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/cost_dif.eps}}
	\vspace*{2pt}
	\caption{Communicative cost by generation under Transmission-only (blue) and Transmission + Communication (red). The upper panels show each individual transmission chain (thin lines; with jitter) and the mean across chains (thick lines). The lower panels show a polynomial regression model of the data (posterior mean and 95\% HDI). Generation~0 is not included in the model, since it does not represent participant data.}
	\label{cost_dif}
	\end{figure}

Fig.~\ref{cost_dif} plots communicative cost over the nine generations of cultural transmission in each condition. In the upper panels, each thin line represents an individual transmission chain, and the thick lines shows the mean across chains. The initial, generation-0 seed systems are all fully informative (zero cost) because each meaning is signaled by a unique arbitrary spelling. In both conditions, cost increases with generation, implying that the languages become less informative over time. This is to be expected in the Transmission-only condition, since there is only pressure from learnability, so the systems evolve to become simple and transparent. In contrast, this is not the expected result in Transmission + Communication, where the communicative pressure should promote the emergence of informative systems. This result does, however, align with our qualitative interpretation of the data, which suggested that participants in the communicative condition rarely produced differentiated spellings, be it through explicit innovation or implicit generalization. Interestingly, however, cost plateaus at around 1.3~bits in Transmission-only and around 1~bit in Transmission + Communication, suggesting that there is some tendency for the systems to remain slightly more informative in the communicative condition.

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/posterior_dif.eps}}
	\vspace*{2pt}
	\caption{Posterior parameter estimates for the Experiment~1 polynomial regression model. Blue curves show results for the Transmission-only condition and red curves show results for the Transmission + Communication condition. $\alpha$ represents the intercept of the polynomial (Generation~5) and the $\beta$ parameters characterize its shape.}
	\label{posterior_dif}
	\end{figure}

To evaluate this argument statistically, we regressed cost on generation in a Bayesian, hierarchical, polynomial regression model. The posterior parameter estimates are shown in Fig.~\ref{posterior_dif}. Essentially, the $\beta$ parameters of the model (characterizing the shape of the polynomial) do not differ substantially between conditions, but the intercepts ($\alpha$) do appear to differ. Since generation was centered (for the purpose of fitting the model), the $\alpha$ parameters correspond to Generation~5 and therefore represent the main plateaus of the polynomials. The posterior difference between $\alpha_\mathrm{tran}$ and $\alpha_\mathrm{comm}$ was 0.301~bits (95\%~HDI: 0.014--0.597). This difference, and its credible interval, is greater than zero, pointing to an effect of condition on the level at which cost plateaus. We emphasize, however, that this difference represents a relatively small effect size, corresponding to a 9.4 percentage point advantage for systems that emerged in the communicative condition (i.e., systems that emerged under Transmission-only would, on average, yield a communicative success rate of 40.7\%, whereas the equivalent value for Transmission + Communication is 50.1\%). Thus, although the presence of the communicative pressure does appear to make a difference to the informativeness of the emergent systems, that difference is not especially impressive considering that an ideal, expressive system of suffix spellings would support a 100\% success rate.

@@

@@

As in Experiment~1, we plot communicative cost by generation in both conditions (see Fig.~\ref{cost_con}). Each colored line represents an individual chain, and the thick black lines shows the mean across chains. These mean lines reveal a fairly clear effect from epoch. During Epoch~I, there is a shallow increase in cost with generation in both conditions. This represents a kind of baseline level of noise when the spoken languages are fully informative. During Epoch~II, there is a steep rise in cost in the Transmission-only condition, but a relatively shallow rise in Transmission + Communication. This is because the communicative pressure acts to slow down the rate at which the systems become transparent or redundant. In Epoch~III, the cost curve plateaus in the Transmission-only condition because the written languages have become sufficiently simple to be reliably learned and reproduced with little change. In Transmission + Communication, by contrast, the curve gets a little steeper because the systems still have some room to simplify and the pressure to simplify has been ramped up.

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/cost_con.eps}}
	\vspace*{2pt}
	\caption{Communicative cost by generation under Transmission-only (blue) and Transmission + Communication (red). The upper panels show each individual transmission chain (thin lines; with jitter) and the mean across chains (thick lines). The lower panels show a polynomial regression model of the data (posterior mean and 95\% HDI). Generation~0 is not included in the model, since it does not represent participant data.}
	\label{cost_con}
	\end{figure}
